### 2019-04-01 2-Chapters/4-Chapter/CrownCom_17.git/IoT_slotted.tex
Updated bibliography for one reference.
Add Figures for Appendix of Chapter4
Done for now for Chapter4

As explained before, setting up the future Internet of Things (IoT) networks will require to support more and more communicating devices.
In this section, we show that intelligent devices in unlicensed bands can use Multi-Armed Bandit (MAB) learning algorithms to improve resource exploitation.
%
We evaluate the performance of two classical MAB learning algorithms, \UCB{} and Thomson Sampling, to handle the decentralized decision-making of Spectrum Access, applied to IoT networks. We also evaluate the learning performance when the number of intelligent end-devices grows.
%
We show that using learning algorithms does help to fit more devices in such networks, even when all end-devices are intelligent and are dynamically switching from channel to channel.
In the studied scenario, stochastic MAB learning provides an up to $16\%$ gain in term of successful transmission probabilities, and has near optimal performance even in non-stationary settings with a majority of intelligent devices.

@@

The aim of this section is to assess the potential gain of learning algorithms in IoT scenarios, even when the number of intelligent devices in the network increases,
and the network is less and less stationary.
% and the stochastic hypothesis is more and more violated.
To do that, we suppose an IoT network made of two types of devices: \emph{static devices} that use only one channel (fixed in time), and \emph{dynamic devices} that can choose the channel for each of their transmissions. Static devices form an interfering traffic, which could have been generated by devices using other standards as well.
Note that instead of assuming that each static device uses a fixed channel, we could also assume a looser hypothesis: if each static device uses a fixed sub-set of the $K$ channels, and a purely uniform random access in its set of considered channels, then \emph{in average} the observed occupancy of the $K$ channels can be modeled as if it were occupied by (more) static devices using only one channel. So our first hypothesis is actually not constraining.
We first evaluate the probability of collision if dynamic devices randomly select channels (that is, a naive approach), and if a centralized controller optimally distributes all of them in the channels at the beginning of the scenario (that is, an ideal approach).
Then, these three reference scenarios allow to evaluate the performance of \UCB{} and TS algorithms in a decentralized network, in terms of successful communication rate, as it reflects the network efficiency.
We show that these algorithms have near-optimal performance, even when the proportion of end-devices increases and the interfering traffic from other devices becomes less and less stochastic.

@@

\paragraph{One gateway and many devices}
%
We consider the system model presented in Figure~\ref{fig:41:system_model1}, where a set of objects all sends uplink packets to the (unique) network gateway.
Messages can be sent in a fixed number $K\geq2$ of wireless channels (\eg, $K=4$ or $K=16$ for instance), which are centered indifferently around the $433.5\;\mathrm{MHz}$ or the $877\;\mathrm{MHz}$ ISM bands (as it is chosen for the implementation presented in Section~\ref{sec:4:gnuradio}).
The communication between IoT devices and this gateway is done through a simple pure ALOHA-based protocol where devices transmit uplink packets of fixed duration whenever they want.
%
Note that this choice of number of channels and of frequency bands is for instance the one of the current LoRa standard in Europe, but of course the precise choice does not really matter, and future standards can choose different parameters.

\begin{figure}[!b]
    \centering
    \includegraphics[width=0.70\linewidth]{system_model1.eps}
    \caption{In our system model, some dynamic devices (in the \textcolor{blue}{IoT network in blue}) transmit packets to a gateway and suffer from the interference generated by neighboring networks (in \textcolor{orange}{orange}).}
    \label{fig:41:system_model1}
\end{figure}

The devices can transmit their packets in one of the $K$ channels. In the case where the gateway --of the corresponding IoT network-- receives an uplink message in one channel, it transmits an acknowledgement to the end-device in the same channel, after a fixed delay (of $1\;\mathrm{s}$ in our implementation).
In a more general setting, if there were more than one gateway in the same IoT network being considered, the network would reply to the object by letting only one gateway sends back an acknowledgement, and the natural choice is for the network to send this acknowledgement by the gateway which received the uplink message.
For simplicity but without loss of generality, we consider a network with only one gateway,
but still it is important to observe the distinction between the gateway, which is simply a RF node, and the IoT network in charge of receiving, handling, and replying to the incoming uplink messages from the IoT objects being paired in the network.
%

These communications operate in unlicensed ISM bands and, consequently, they can suffer from interference generated by uncoordinated neighboring networks (in \textcolor{orange}{orange} in Figure~\ref{fig:41:system_model1}).
This interfering traffic is uncontrolled, and is most likely unevenly distributed over the $K$ different channels.
%
To simulate networks designed for the Internet of Things (IoT), we consider a protocol \textbf{with no sensing} and no repetition of uplink messages. The gateway is in charge of sending back an acknowledgement, after some fixed-time delay (typically $1\;\mathrm{s}$), to any object paired with this gateway in this network (\ie, in \textcolor{blue}{blue} in Figure~\ref{fig:41:system_model1}) who succeeded in sending an uplink packet.
%
By considering a small number of wireless channels, and a unique PHY layer configuration (\ie, modulation, waveform, etc), and in case of a non-uniform traffic in the different channels,
the object can improve their usage of the network if they are able to \emph{learn} on the fly the best channels to use, that is, the most vacant one. Indeed, in this model, the \emph{quality} of the channels (\ie, the ressources) are identified by their vacancy, or one minus their occupancy rates.

@@

\begin{figure}[!t]
    \centering
    \includegraphics[scale=0.40]{protocol.eps}
    \caption{The considered time-frequency slotted protocol. Each frame is composed by a fix-duration \textcolor{blue}{uplink slot} in which the end-devices transmit their \textcolor{blue}{(uplink) packets}. If a packet is well received, the gateway replies by transmitting an \textcolor{darkgreen}{\Ack}, after the ack delay.}
    \label{fig:41:protocol}
\end{figure}

\paragraph{Static vs dynamic devices}
%
We make the following assumptions on the network.
We assume that there are two types of end-devices in the network:
\begin{itemize}
    \item
    \emph{Static} end-devices are identical, and each of them uses only one channel to communicate with the gateway. Their choice is assumed to be fixed in time (\ie, stationary). The traffic generated by these devices is considered as an interfering traffic for other devices.
    \item
    \emph{Dynamic} (or \emph{smart}) end-devices can use all the available channels, by quickly changing their communication channel at any time. They can also store communication successes or failures they experienced in each channel, in order to change channel, possibly at every time slot.
    We also assume that the dynamic end-devices can run simple embedded decision making algorithms, and have limited but reasonable computing as well as storage capacities.
    Note that of course we assume a limited storage capacity, so no object stores the full history of communication successes and failures, but they only store average rates (which can be stored using one $32$-bits IEEE float number for each of the $K$ channels).
    Our goal is to propose a learning algorithm than can be implemented by each dynamic devices, in a decentralized and independent manner, to improve their successful communication rate automatically.
\end{itemize}

@@

\paragraph{Random emission patterns}
%
We suppose that all devices follow the same emission pattern, being fixed in time, and we choose to model it as a simple Bernoulli process:
all devices have the same probability to send a packet in any (discrete) temporal slot, and we denote $p \in (0, 1)$ this probability\footnote{~In the experiments below, $p$ is about $10^{-3}$, because in a crowded network $p$ should be smaller than $K / (S + D)$ for all devices to communicate successfully (in average).}.
The parameter $p$ essentially controls the frequency of communication for each device, once the time scale is fixed (\ie, real time during two messages), and $1/p$ is proportional to the \emph{duty cycle}.
For instance for IoT objects sending messages with a duration of $1$ second exactly, a ``smart'' refrigerator could have to send a daily message, and if a message can be sent every second, then $p = 1 / (12 \times 60 \times 60) \simeq 1.5 \times 10^{-5}$.

\paragraph{Some assumptions on the network occupation}
%
We focus on \emph{dense networks}, in which the number of devices $S + D$ is very large compared to $K$ (for instance, about $1000$ to $100000$, while $K$ is about $4$ to $256$).
As this problem is only interesting if devices are able to communicate reasonably efficiently with the gateway, we assume that devices only communicate occasionally, \ie, with a low \emph{duty cycle}, as it is always considered for IoT.
Note that even unlicensed bands have such limitations, as for instance this is a strict requirement for any object using the $867\;\mathrm{MHz}$ ISM band (and its regulation also enforces a maximum transmission power, but we do not consider power transmission in this work).
%
We prefer this choice rather than non-crowded networks, \ie, where $S + D \leq K$, as the former makes more sense for IoT networks.
% \footnote{~Here, $K$ will typically be about $10$ different radio channels, usually in the same RF band and with separate mean carrier frequencies, and the number of devices will be of the order of $1000$ to $10000$.}.
Indeed, imagine if there were only $K=4$ channels being occupied by $D+S = 100$ objects, each communicating with a high rate of $p=1/10$, with a non-zero occupancy in each channel (\ie, $\forall i, S_i > 0$). Then, almost all time slots will lead to collisions in all channels, for a uniformly random access scheme, and thus the network efficiency (\ie, successful communication rate) will be so close to zero that using learning algorithms cannot improve much.
Such scenario is not our target of interest, and thus we prefer to only consider feasible scenarios where $p$ is smaller than $K/(S+D)$ to have an average of active objects not larger than the number of channels.

\paragraph{MAB model}
%
We consider the network from the point of view of one dynamic device.
Following a Bernoulli emission process, every times a dynamic device has to communicate with the gateway,
it has to choose one channel (at each transmission $t \geq 1, t \in \mathbb{N}$), denoted as $C(t) = k \in\{1,\dots,K\}$.
Then, the dynamic device waits in this channel $C(t)$ for an acknowledgement sent by the gateway, during a certain period (\eg, $5$ seconds for the example of a ``smart fridge'' considered above).
Before sending another message (\ie, at time $t+1$), the dynamic device knows if it received or not this \Ack{} message.
%
For this reason, selecting channel (or arm) $k$ at time $t$ yields a (random) feedback, called a (binary) \emph{reward}, $r_k(t) \in \{0,1\}$, being $0$ if no \Ack{} was received before the next message, or $1$ if \Ack{} was successfully received.
The goal of the dynamic device is to minimize its packet loss ratio, or equivalently, to maximize its number of successful transmissions, or its cumulative reward,
$r_{1 \dots T} := \sum_{t = 1}^T r_{C(t)}(t),$
which is a usual objective in MAB problems.

@@

The problem parameters
% are $\mu_1,\dots,\mu_K$, the mean availability of the channels,
are $S_1, \dots, S_K$ the (stationary) occupancy of the channels by the static devices,
and they are of course unknown to each dynamic device, so to maximize their cumulated rewards, they must learn the distributions of the channels, in order to be able to progressively improve its successful communication rate.
%
The goal is thus to propose a simple sequential algorithm to be applied identically and independently by each dynamic device, in a fully distributed setting (each device runs its own algorithm, from its observations), in order to minimize collisions and maximize the fraction of successful transmissions of all the dynamic devices.
%
This requires to tackle the so-called \emph{exploration-exploitation dilemma}: a device has to try all arms a sufficient number of times to get a robust estimate of their qualities, while not selecting the worst arms too many times.

@@

We investigate here the optimal policy that can be achieved if the dynamic devices have a perfect knowledge of $(S_i)_i$ the repartition of static devices, and a fully centralized decision making\footnote{~This optimal policy needs an \emph{oracle} seeing the entire system, and affecting all the dynamic devices, once and for all, for a fixed stationary scenario, in order to avoid any signaling overhead.} is possible.
We want to find the stationary repartition of devices into channels that maximizes the probability of having a successful transmission.

@@

\begin{proof}
\begin{itemize}
    \item
    In a realistic scenario, we can assume that $D_i\leq \frac{-2}{\ln\left(1-p\right)} \approx \frac{2}{p},\quad \forall i\in\llbracket 1;K \rrbracket$. For such values for $D_i$, the objective function $f: (D_1, \dots, D_{K}) \mapsto \sum_{i=1}^{K} D_i (1 - p)^{S_i + D_i -1}$ is concave as the sum of concave functions
    \footnote{~It worth noting that $f$ is neither concave nor quasi-concave on $[0,\infty)^{K}$ \cite{Luenberger68,Yaari77}.}.
    \item
    The Lagrange multipliers method can be applied to the optimization problem \eqref{eq:41:optPb}, with a concave objective function $f$, linear equality constraints \eqref{eq:41:eqCstr} and linear inequality constraints \eqref{eq:41:ineqCstr}. The strong duality condition is satisfied in this case \cite{BoydVanderberghe04}, so finding the saddle points will be enough to find the maximizers.
    \item
    More details are given in Section~\ref{sec:5:proofLagrangian} in the Appendix of this Chapter.
    % \hfill{}$\square$
    \end{itemize}
\end{proof}

@@

While the stochastic MAB model has been used to describe some aspects of Cognitive Radio systems, it is in principle not suitable for our IoT model, due to the non-stationarity of the channels occupancy caused partly by the learning policy used by dynamic objects, but mainly by their random activation processes.

In our model, every dynamic device implements its own learning algorithm, \emph{independently}.
For one device, the time $t$ refers to the number of time it accessed the network (following its Bernoulli transmission process, \ie, its duty cycle), \emph{not} the total number of time slots from the beginning, as rewards are only obtained after a transmission, and IoT objects only transmit sporadically, due to low transmission duty cycles.

@@

% Considered alone, each dynamic device
We propose algorithms in which each dynamic device ignores all the other one and
implements a learning algorithm to play a bandit game.
%
In each time slot, if it has to communicate (which happens with probability $p$), then it chooses a channel and it receives a reward $1$ if the transmission is successful, $0$ otherwise.
Each device aims at maximizing the sum of the rewards collected during its communication instants, which shall indeed maximize the fraction of successful transmissions. Besides the modified time scale (rewards are no longer collected at every time step), this looks like a usual bandit problem.
However, it cannot be modeled as a stochastic MAB, as the rewards are clearly \emph{not} \iid: they not only depend on the (stationary, \iid) behavior of the static devices, but also on the behavior of other dynamic devices, that is not stationary (because of learning and random activation of each devices).
%
Despite this, we show in the next subsection that running a stochastic bandit algorithm for each device based on its own rewards is surprisingly successful.

@@

The two algorithms, \UCB{} and Thompson sampling (TS), were presented in Section~\ref{sec:2:famousMABalgorithms}.
The \UCB{} algorithm uses a parameter $\alpha > 0$, set to $\alpha = 1/2$ in our experiments.
% originally $\alpha$ was set to $2$ \cite{Auer02}, but empirically $\alpha = 1/2$ is known to work better (uniformly across problems), and $\alpha > 1/2$ is advised by the theory \cite{Bubeck12}.
%
In our model, every dynamic dynamic device implements its own \UCB{} algorithm, \emph{independently}.
For one dynamic device, the time $t$ is the total number of sent messages from the beginning, as rewards are only obtained after a transmission.
%
The TS algorithm, although being simple and easy to implement, is known to perform well for stochastic problems, for which it was proven to be asymptotically optimal \cite{AgrawalGoyal11,Kaufmann12Thompson}.
It is known to be empirically efficient, and for these reasons it has been used successfully in various applications, including on problems from Cognitive Radio \cite{Toldov16,Mitton16}, and also in previous work on decentralized IoT-like networks \cite{Darak16}.

@@

Instead of using MAB algorithms assuming a stochastic hypothesis on the system, we could try to use MAB algorithms designed to tackle a more general problem, that makes no hypothesis on the interfering traffic.
The \emph{adversarial MAB} algorithms is a broader family, of which a well-known and efficient example is the $\mathrm{Exp}3$ algorithm \cite{Bubeck12}.
Empirically, the $\mathrm{Exp}3$ algorithm turned out to perform significantly worse than both \UCB{} and TS in the same experiments,
therefore we did not report results here.
% FIXME maybe we should... ?
%(see Section \ref{sub:41:numericalResults}).
However, contrarily to the two stochastic algorithms, the use of $\mathrm{Exp}3$ is correctly justified, even in the non-stationary model, as its performance guarantees are true in \emph{any} setting.
% But it is not so surprising that it performs worse, as the theoretical performance guarantees of adversarial MAB algorithms are an order of magnitude
% %\footnote{~We preferred not to talk about \emph{regret} in this study, but in a few words: it is a measure of how \emph{bad} the algorithm performed in terms of its accumulated rewards, in comparison to the best possible policy, and should be as small as possible. For stochastic algorithms, being ``efficient'' means having a regret bounded as $R_T = \mathcal{O}(\log T)$, but for adversarial algorithms, it means having $R_T = \mathcal{O}(\sqrt{K T})$.}
% worse than the one for stochastic ones.
% (in their respective case of application).
Further research on this aspect could lead to interesting future works.

% ------------------------------------------------------------------------
\subsection{Numerical results on an example of IoT network}\label{sub:41:numericalResults}

@@

\begin{figure}[!t]
    \centering
    \subfloat[10\% of dynamic devices]{\includegraphics[scale=0.5]{10intelligent.eps}
    \label{fig:41:10intelligent}}
    \hfill
    \subfloat[30\% of dynamic devices]{\includegraphics[scale=0.5]{30intelligent.eps}
    \label{fig:41:30intelligent}}
    \vskip\baselineskip
    \vspace*{-20pt}
    \subfloat[50\% of dynamic devices]{\includegraphics[scale=0.5]{50intelligent.eps}
    \label{fig:41:50intelligent}}
    \hfill
    \subfloat[100\% of dynamic devices]{\includegraphics[scale=0.5]{100intelligent.eps}
    \label{fig:41:100intelligent}}
    \caption{Performance of two MAB algorithms (\UCB{} and Thompson Sampling), compared to extreme references without learning or oracle knowledge, when the proportion of dynamic end-devices in the network increases, from $10\%$ to $100\%$.}
    \label{fig:41:from10to100}
    \vspace*{-10pt}
\end{figure}

@@

>>>>

As explained before, the future Internet of Things (IoT) networks will require to support more and more communicating devices.
In this section, we show that intelligent devices in unlicensed bands can use Multi-Armed Bandit (MAB) learning algorithms to improve spectral resource exploitation.
%
We evaluate the performance of two classical MAB learning algorithms, \UCB{} and Thomson Sampling, to handle the decentralized decision-making of Spectrum Access, applied to IoT networks.
We mean by \emph{decentralized} that learning is performed on the device side, and is spread on the totality of the devices in a network.
We also evaluate the learning performance when the number of intelligent end-devices grows.
%
We show that using learning algorithms
firstly extends devices battery autonomy (by reducing the packet-loss-ratio or equivalently, by improving the successful-communication rate),
and secondly does help to fit more devices in such networks, even when all end-devices are intelligent and are dynamically switching from channel to channel.
In the studied settings,
Stochastic MAB learning
% provides an up to $16\%$ gain in terms of successful transmission probabilities, and
has near optimal performance even in non-stationary settings with a majority of intelligent devices.

@@

The aim of this section is to assess the potential gain of learning algorithms in IoT scenarios, even when the number of intelligent devices in the network increases,
and the network is less and less stationary.
% and the stochastic hypothesis is more and more violated.
To do that, we suppose an IoT network made of two types of devices: \emph{static devices} that use only one channel (fixed in time), and \emph{dynamic devices} that can choose the channel for each of their transmissions. Static devices form an interfering traffic, which could be generated by devices using other standards as well.
Note that instead of assuming that each static device uses a fixed channel, we could also assume a looser hypothesis: if each static device uses a fixed sub-set of the $K$ channels, and a purely uniform random access in its set of considered channels, then \emph{in average} the observed occupancy of the $K$ channels can be modeled as if it were occupied by (more) static devices using only one channel. So our first hypothesis is actually not constraining.
We first evaluate the probability of collision if dynamic devices randomly select channels (that is, a naive approach), and if a centralized controller would optimally distribute all of them in the channels at the beginning of the scenario.
This second approach is ideal, but not realistic in decentralized IoT networks, and it is just used here as a reference.
Then, these three reference scenarios allow to evaluate the performance of \UCB{} and TS algorithms in a decentralized network, in terms of successful communication rate, as it reflects the network efficiency.
We show that these algorithms have near-optimal performance, even when the proportion of end-devices increases and the interfering traffic from other devices becomes less and less stochastic.

@@

\paragraph{One gateway and many devices}
%
We consider the system model presented in Figure~\ref{fig:41:system_model1}, where a set of devices all sends uplink packets to a (unique) network gateway.
Messages can be sent in a fixed number $K\geq2$ of wireless channels,
that are assumed to be orthogonal, that is, a message in channel $i$ will not cause collision to a message in another channel $j \neq i$.
% (\eg, $K=4$ or $K=16$ for instance), which are centered indifferently around the $433.5\;\mathrm{MHz}$ or the $877\;\mathrm{MHz}$ ISM bands (as it is chosen for the implementation presented in Section~\ref{sec:4:gnuradio}).
The communication between IoT devices and this gateway is done through a simple pure ALOHA-based protocol where devices transmit uplink packets of fixed duration whenever they want.
%
% Note that this choice of number of channels and of frequency bands is for instance the one of the current LoRa standard in Europe, but of course the precise choice does not really matter, and future standards can choose different parameters.

\begin{figure}[!t]
    \centering
    \includegraphics[width=0.70\linewidth]{system_model1.eps}
    \caption{In our system model, some dynamic devices (in the \textcolor{blue}{IoT network in blue}) transmit packets to a gateway and suffer from the interference generated by neighboring networks (in \textcolor{orange}{orange}).}
    \label{fig:41:system_model1}
\end{figure}

The devices can transmit their packets in one of the $K$ channels. In the case where the gateway --of the corresponding IoT network-- receives an uplink message in one channel, it transmits an acknowledgement to the end-device in the same channel, after a fixed delay.
This is realistic in IoT networks such as networks based on LoRa.
% (of $1\;\mathrm{s}$ in our implementation).
In a more general setting, if there were more than one gateway in the same IoT network being considered, the network would reply to the device by letting only one gateway sends back an acknowledgement.
The natural choice is for the network to send this acknowledgement by the gateway which received the uplink message.
For simplicity but without loss of generality, we consider a network with only one gateway,
but still it is important to observe the distinction between the gateway, which is simply a RF node, and the IoT network in charge of receiving, handling, and replying to the incoming uplink messages from the IoT devices being paired in the network.
%

These communications operate in unlicensed ISM bands and, consequently, they can suffer from interference generated by uncoordinated co-localized and/or neighboring networks (in \textcolor{orange}{orange} in Figure~\ref{fig:41:system_model1}).
This interfering traffic is uncontrolled, and is most likely unevenly distributed over the $K$ different channels,
as each IoT protocol or IoT networks can choose to use a different subset of channels.
%
In order to simulate networks designed for the IoT, we consider a protocol \textbf{with no sensing} and no repetition of uplink messages. The gateway is in charge of sending back an acknowledgement, after some fixed-time delay (typically $1\;\mathrm{s}$), to any device paired with this gateway in this network (\ie, in \textcolor{blue}{blue} in Figure~\ref{fig:41:system_model1}) who succeeded in sending an uplink packet.
%
By considering a small number of wireless channels, and a unique PHY layer configuration\footnote{~But MAB learning to also optimize on the PHY layer parameters have started to be explored, after our work \cite{Bonnefoi17}, see for instance \cite{KerkoucheAlami18}.} (\ie, modulation, waveform, etc), and in case of a non-uniform traffic in the different channels,
the device can improve their usage of the network if they are able to \emph{learn} on the fly the best channels to use, that is, the most vacant one. Indeed, in this model, the \emph{quality} of the channels (\ie, the ressources) are identified by their vacancy, or one minus their occupancy rates.

@@

\begin{figure}[!t]
    \centering
    \includegraphics[scale=0.40]{protocol.eps}
    \caption{The considered time-frequency slotted protocol. Each frame is composed by a fixed duration \textcolor{blue}{uplink slot} in which the end-devices transmit their \textcolor{blue}{(uplink) packets}. If a packet is well received, the gateway replies by transmitting an \textcolor{darkgreen}{\Ack}, after the ack delay.}
    \label{fig:41:protocol}
\end{figure}

\paragraph{Static vs dynamic devices}
%
We make the following assumptions on the network.
We assume that there are two types of end-devices in the network:
\begin{itemize}
    \item
    \emph{Static} end-devices are identical, and each of them uses only \emph{one} channel to communicate with the gateway, with no loss of generality.
    %  if devices random access the channels.
    Their choice is assumed to be fixed in time (\ie, stationary). The traffic generated by these devices is considered as an interfering traffic for other devices.
    \item
    \emph{Dynamic} (or \emph{smart}) end-devices can use all the available channels, by quickly changing their communication channel at any time slot, following a Machine Learning policy.
    For that end, they use their history of past communication successes or failures they experienced in each channel, in order to learn about channel availability.
    We also assume that the dynamic end-devices can run simple embedded decision making algorithms, and have limited but reasonable computing as well as storage capacities.
    Note that of course we assume a limited storage capacity, so no device stores the full history of communication successes and failures, but they only store average rates (which can be stored using one $32$-bits IEEE float number for each of the $K$ channels).
    Our goal is to propose a learning algorithm than can be implemented by each dynamic device, in a decentralized and independent manner, to improve their successful communication rate automatically.
\end{itemize}

@@

\paragraph{Random emission patterns}
%
We suppose that all devices follow the same emission pattern, being fixed in time, and we choose to model it as a simple Bernoulli process:
all devices have the same probability to send a packet in any (discrete) temporal slot, and we denote $p \in (0, 1)$ this probability\footnote{~In the experiments below, $p$ is about $10^{-3}$, because in a crowded network $p$ should be smaller than $K / (S + D)$ for all devices to communicate successfully (in average).}.
The parameter $p$ essentially controls the frequency of communication for each device, once the time scale is fixed, and $1/p$ is proportional to the \emph{duty cycle}.
% (\ie, real time during two messages)
For instance, for IoT devices sending messages with a duration of $1$ second and possibly at every second, then on a daily basis we have $p = 1 / (12 \times 60 \times 60) \simeq 1.5 \times 10^{-5}$.

\paragraph{Some assumptions on the network occupation}
%
We focus on \emph{dense networks}, in which the number of devices $S + D$ is very large compared to $K$ (for instance, about $1000$ to $100000$, while $K$ is about $4$ to $256$).
As this problem is only interesting if devices are able to communicate reasonably efficiently with the gateway, we assume that devices only communicate occasionally, \ie, with a low \emph{duty cycle}, as it is always considered for IoT.
Note that even unlicensed bands have such limitations, as for instance this is a strict requirement for any device using the $867\;\mathrm{MHz}$ ISM band in Europe (and its regulation also enforces a maximum transmission power, but we do not consider power transmission in this work).
%
We prefer this choice rather than non-crowded networks, \ie, where $S + D \leq K$, as the former makes more sense for IoT networks.
% \footnote{~Here, $K$ will typically be about $10$ different radio channels, usually in the same RF band and with separate mean carrier frequencies, and the number of devices will be of the order of $1000$ to $10000$.}.
Indeed, imagine if there were only $K=4$ channels being occupied by $D+S = 100$ devices, each communicating with a high rate of $p=1/10$, with a non-zero occupancy in each channel (\ie, $\forall i, S_i > 0$). Then, almost all time slots will lead to collisions in all channels, for a uniformly random access scheme, and thus the network efficiency (\ie, successful communication rate) will be so close to zero that using learning algorithms cannot improve much.
Such scenario is not our target of interest, and thus we prefer to only consider feasible scenarios where $p$ is smaller than $K/(S+D)$, in order to have an average of active devices not larger than the number of channels.

\paragraph{MAB model}
%
We consider the network from the point of view of one dynamic device.
Following a Bernoulli emission process, every times a dynamic device has to communicate with the gateway,
it has to choose one channel (at each transmission $t \geq 1, t \in \mathbb{N}$), denoted as $C(t) = k \in\{1,\dots,K\}$.
Then, the dynamic device waits in this channel $C(t)$ for an acknowledgement sent by the gateway, during a certain period (\eg, $5$ seconds for the example considered above).
Before sending another message (\ie, at time $t+1$), the dynamic device knows if it received or not this \Ack{} message.
%
For this reason, selecting channel (or arm) $k$ at time $t$ yields a (random) feedback, called a (binary) \emph{reward}, $r_k(t) \in \{0,1\}$, being $0$ if no \Ack{} was received before the next message, or $1$ if \Ack{} was successfully received.
The goal of the dynamic device is to minimize its packet loss ratio, or equivalently, to maximize its number of successful transmissions, or its cumulative reward,
$r_{1 \dots T} := \sum_{t = 1}^T r_{C(t)}(t),$
which is a usual objective in MAB problems.

@@

The problem parameters
% are $\mu_1,\dots,\mu_K$, the mean availability of the channels,
are $S_1, \dots, S_K$ the (stationary) occupancy of the channels by the static devices,
and they are unknown to each dynamic device, so to maximize their cumulated rewards, they must learn the distributions of the channels, in order to be able to progressively improve their respective successful communication rate.
%
The goal is thus to propose a simple sequential algorithm to be applied identically and independently by each dynamic device, in a fully distributed setting (each device runs its own algorithm, from its observations), in order to minimize collisions and maximize the fraction of successful transmissions of all the dynamic devices.
%
This requires to tackle the so-called \emph{exploration-exploitation dilemma}: a device has to try all arms a sufficient number of times to get a robust estimate of their qualities, while not selecting the worst arms too many times.

@@

We investigate here the optimal policy that can be achieved if the dynamic devices have a perfect knowledge of $(S_i)_i$ the repartition of static devices, and a fully centralized decision making\footnote{~This optimal policy needs an \emph{oracle} seeing the entire system, and affecting all the dynamic devices, once and for all, for a fixed stationary scenario, in order to avoid any signaling overhead. This is not possible in IoT contexts as several completely independent networks may operate in a single place.} is possible.
We want to find the stationary repartition of devices into channels that maximizes the probability of having a successful transmission.

@@

\begin{proof}
\begin{itemize}
    \item
    In a realistic scenario, we can assume that $D_i\leq \frac{-2}{\ln\left(1-p\right)} \approx \frac{2}{p},\quad \forall i\in\llbracket 1;K \rrbracket$. For such values for $D_i$, the objective function $f: (D_1, \dots, D_{K}) \mapsto \sum_{i=1}^{K} D_i (1 - p)^{S_i + D_i -1}$ is concave as the sum of concave functions
    \footnote{~It is worth noting that $f$ is neither concave nor quasi-concave on $[0,\infty)^{K}$ \cite{Luenberger68,Yaari77}.}.
    \item
    The Lagrange multipliers method can be applied to the optimization problem \eqref{eq:41:optPb}, with a concave objective function $f$, linear equality constraints \eqref{eq:41:eqCstr} and linear inequality constraints \eqref{eq:41:ineqCstr}. The strong duality condition is satisfied in this case \cite{BoydVanderberghe04}, so finding the saddle points will be enough to find the maximizers.
    \item
    More details are given in Section~\ref{sec:4:proofLagrangian} in the Appendix of this Chapter.
    % \hfill{}$\square$
    \end{itemize}
\end{proof}

@@

While the stochastic MAB model has been used to describe some aspects of Cognitive Radio systems, it is in principle not suitable for our IoT model, due to the non-stationarity of the channels occupancy caused partly by the learning policy used by dynamic devices, but mainly by their random activation processes.

In our model, every dynamic device implements its own learning algorithm, \emph{independently}.
For one device, the time $t$ refers to the number of time it accessed the network (following its Bernoulli transmission process, \ie, its duty cycle), \emph{not} the total number of time slots from the beginning, as rewards are only obtained after a transmission, and IoT devices only transmit sporadically, due to low transmission duty cycles.

@@

% Considered alone, each dynamic device
We propose algorithms in which each dynamic device ignores all the other one and
implements a learning algorithm to play a bandit game.
%
In each time slot, if it has to communicate (which happens with probability $p$), then it chooses a channel and it receives a reward $1$ if the transmission is successful, $0$ otherwise.
Each device aims at maximizing the sum of the rewards collected during its communication instants, which shall indeed maximize the fraction of successful transmissions. Besides the modified time scale (rewards are no longer collected at every time step), this looks like a usual bandit problem.
However, it cannot be modeled as a stochastic MAB, as the rewards are (unfortunately) \emph{not} \iid: they not only depend on the (stationary, \iid) behavior of the static devices, but also on the behavior of other dynamic devices, that is not stationary (because of learning and random activation of each devices).
%
Despite this, we show in the next subsection that running a stochastic bandit algorithm for each device based on its own rewards is surprisingly successful.

@@

The two algorithms, \UCB{} and Thompson sampling (TS), were presented in Section~\ref{sec:2:famousMABalgorithms}.
The \UCB{} algorithm uses a parameter $\alpha > 0$, set to $\alpha = 1/2$ in our experiments.
% originally $\alpha$ was set to $2$ \cite{Auer02}, but empirically $\alpha = 1/2$ is known to work better (uniformly across problems), and $\alpha > 1/2$ is advised by the theory \cite{Bubeck12}.
%
In our model, every dynamic device implements its own \UCB{} algorithm, \emph{independently}.
For one dynamic device, the time $t$ is the total number of sent messages from the beginning, as rewards are only obtained after a transmission.
%
The TS algorithm, although being simple and easy to implement, is known to perform well for stochastic problems, for which it was proven to be asymptotically optimal \cite{AgrawalGoyal11,Kaufmann12Thompson}.
It is known to be empirically efficient, and for these reasons it has been used successfully in various applications, including problems from Cognitive Radio \cite{Toldov16,Mitton16}, and also in previous work on decentralized IoT-like networks \cite{Darak16}.

@@

Instead of using MAB algorithms assuming a stochastic hypothesis on the system, we could try to use MAB algorithms designed to tackle a more general problem, that makes no hypothesis on the interfering traffic.
The \emph{adversarial MAB} algorithms is a broader family, of which a well-known and efficient example is the $\mathrm{Exp}3$ algorithm \cite{Auer02,Bubeck12}.
Empirically, the $\mathrm{Exp}3$ algorithm turned out to perform significantly worse than both \UCB{} and TS in the same experiments,
therefore we did not report results here.
% FIXME maybe we should... ?
%(see Section \ref{sub:41:numericalResults}).
However, contrarily to the two stochastic algorithms, the use of $\mathrm{Exp}3$ is correctly justified, even in the non-stationary model, as its performance guarantees are true in \emph{any} setting.
% But it is not so surprising that it performs worse, as the theoretical performance guarantees of adversarial MAB algorithms are an order of magnitude
% %\footnote{~We preferred not to talk about \emph{regret} in this study, but in a few words: it is a measure of how \emph{bad} the algorithm performed in terms of its accumulated rewards, in comparison to the best possible policy, and should be as small as possible. For stochastic algorithms, being ``efficient'' means having a regret bounded as $R_T = \mathcal{O}(\log T)$, but for adversarial algorithms, it means having $R_T = \mathcal{O}(\sqrt{K T})$.}
% worse than the one for stochastic ones.
% (in their respective case of application).
Further research on this aspect could lead to interesting future works.

% ------------------------------------------------------------------------
\subsection{Numerical results}\label{sub:41:numericalResults}

@@

\begin{figure}[!b]
    \centering
    \subfloat[10\% of dynamic devices]{\includegraphics[scale=0.5]{10intelligent.eps}
    \label{fig:41:10intelligent}}
    \hfill
    \subfloat[30\% of dynamic devices]{\includegraphics[scale=0.5]{30intelligent.eps}
    \label{fig:41:30intelligent}}
    \vskip\baselineskip
    \vspace*{-20pt}
    \subfloat[50\% of dynamic devices]{\includegraphics[scale=0.5]{50intelligent.eps}
    \label{fig:41:50intelligent}}
    \hfill
    \subfloat[100\% of dynamic devices]{\includegraphics[scale=0.5]{100intelligent.eps}
    \label{fig:41:100intelligent}}
    \caption{Performance of two MAB algorithms (\UCB{} and Thompson Sampling), compared to extreme references without learning or oracle knowledge, when the proportion of dynamic end-devices in the network increases, from $10\%$ to $100\%$.}
    \label{fig:41:from10to100}
    \vspace*{-10pt}
\end{figure}

@@

\begin{figure}[!h]
    \centering
    \includegraphics[scale=0.60]{perf_learning.eps}
    \caption{Learning with \UCB{} and Thomson Sampling, with many dynamic devices. Both algorithms achieve close to optimal performance after a reasonable learning time.}
    \label{fig:41:perf_learning}
\end{figure}
